Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-02-02 causal-bandit/icml-2015/general.tex
updates

\newcommand{\calP}{\mathcal P}
\newcommand{\calA}{\mathcal A}
\newcommand{\x}{\boldsymbol{x}}

We now consider the more general problem where the graph structure is known, but arbitrary.
As before $Y$ is the target variable and $\P{Y=1|X} = r(X)$ is the expected reward given $X$.
We let $\calP_i \subseteq \set{1,\ldots,N}$ denote the parents of variable $X_i$ so the joint distribution over $X$ may be written as
\eq{
\P{X = x} = \prod_{i=1}^N \P{X_i = x_i|X_j = x_j \text{ for all } j \in \calP_i}\,.
}
Throughout this section we assume that the causal structure and conditional distributions of $X$ are known,
which allows us to compute the above quantity for arbitrary $x \in \Omega = \set{0,1}^N$. Given the learner intervens by $do(X_i = j)$
leads to
\eq{
&\Pn{ij}{X = x} = \P{X = x|do(X_i = j)} \\
&\qquad =\prod_{k \neq i} \P{X_k = x_k|X_j = x_j \text{ for all } j \in \calP_k}\,.
}
The expected reward by making the intervention $do(X_i = j)$ is
\eq{
\mu_{ij} = \sum_{x \in \Omega} \P{X = x|do(X_i = j)} r(x)\,.
}
The problem of learning the optimal intervention is made more complicated than the simple case because of 
the dependencies between the random variables. Learning from observational data is no longer straightforward
because variables can be correlated through their parents. Consider, for example, the causual graph below where $X_1$
deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
then intervening on either $X_2$ or $X_3$ is optimal. But the positive reward is never observed from observational samples, despite
the fact that $\P{X_2 = 1} = \P{X_3 = 1} = \frac{1}{2}$.
\begin{figure}[h]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]
\node[main node](1){$X_1$};
\node[main node, below left=of 1](2){$X_2$};
\node[main node, below right=of 1](3){$X_3$};
\node[main node, below right=of 2](4){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (1) edge (3)
    (3) edge (4)
    (2) edge (4);
\end{tikzpicture}
\eq{
\P{X_1 = 1} &= \frac{1}{2} \\
\P{X_2 = j|X_1 = j} &= 
\P{X_3 = j|X_1 = j} = 1\,.
}
\caption{}\label{fig:causalStructure_confounded}
\end{figure}

>>>>

\newcommand{\calP}{\mathcal P}
\newcommand{\calA}{\mathcal A}
\newcommand{\x}{\boldsymbol{x}}
\newcommand{\Ps}{\operatorname{P}}

We now consider the more general problem where the graph structure is known, but arbitrary.
We use the notation $\P{\parents{Y}(X)}$ to denote a random variable that is the probability that the parents of $Y$ are $X$.
Let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ be the mixture distribution.
Let $a \in \calA$ be an intervention and define random variable
\eq{
Z_a(X) = R_a(X) Y \ind{R_a(X) \leq B_a}\,,
}
where $B_a \geq 0$ is some constant to be chosen subsequently and
\eq{
R_a(X) = \frac{\P{\parents{Y}(X)|a}}{\Q{\parents{Y}(X)}}\,.
}
\begin{algorithm}[H]
\begin{algorithmic}
\STATE {\bf Input:} $\eta$, $B$, $T$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $I_t = a$ with probability $\eta$
\STATE Do action $I_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\STATE For each $a \in \calA$ compute an estimate of its return:
\eq{
\forall a\in \calA \qquad \hat \mu_a = \frac{1}{T} \sum_{t=1}^T Z_a(X_t)
}
\STATE Choose $I = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}

\begin{theorem}
Blah
\end{theorem}

As before $Y$ is the target variable and $\P{Y=1|X} = r(X)$ is the expected reward given $X$.
We let $\calP_i \subseteq \set{1,\ldots,N}$ denote the parents of variable $X_i$ so the joint distribution over $X$ may be written as
\eq{
\P{X = x} = \prod_{i=1}^N \P{X_i = x_i|X_j = x_j \text{ for all } j \in \calP_i}\,.
}
Throughout this section we assume that the causal structure and conditional distributions of $X$ are known,
which allows us to compute the above quantity for arbitrary $x \in \Omega = \set{0,1}^N$. Given the learner intervens by $do(X_i = j)$
leads to
\eq{
&\Pn{ij}{X = x} = \P{X = x|do(X_i = j)} \\
&\qquad =\prod_{k \neq i} \P{X_k = x_k|X_j = x_j \text{ for all } j \in \calP_k}\,.
}
The expected reward by making the intervention $do(X_i = j)$ is
\eq{
\mu_{ij} = \sum_{x \in \Omega} \P{X = x|do(X_i = j)} r(x)\,.
}
The problem of learning the optimal intervention is made more complicated than the simple case because of 
the dependencies between the random variables. Learning from observational data is no longer straightforward
because variables can be correlated through their parents. Consider, for example, the causual graph below where $X_1$
deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
then intervening on either $X_2$ or $X_3$ is optimal. But the positive reward is never observed from observational samples, despite
the fact that $\P{X_2 = 1} = \P{X_3 = 1} = \frac{1}{2}$.
\begin{figure}[h]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]
\node[main node](1){$X_1$};
\node[main node, below left=of 1](2){$X_2$};
\node[main node, below right=of 1](3){$X_3$};
\node[main node, below right=of 2](4){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (1) edge (3)
    (3) edge (4)
    (2) edge (4);
\end{tikzpicture}
\eq{
\P{X_1 = 1} &= \frac{1}{2} \\
\P{X_2 = j|X_1 = j} &= 
\P{X_3 = j|X_1 = j} = 1\,.
}
\caption{}\label{fig:causalStructure_confounded}
\end{figure}
